The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-07-18 paper.tex
More Irish analysis

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{fullpage}
\usepackage{natbib}
\usepackage{mathpazo}
\usepackage{tipa}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{linguex}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows}
\usepackage{slashbox}

@@

\ex. \a. /alb\textipa{@}/ $\rightarrow$ [al\textipa{@b@}] \\
         {\it Alba} `Scotland'
     \b. /dorx\textipa{@}/ $\rightarrow$ [dor\textipa{@x@}] \\
         {\it dorcha} `dark'
     \b. /gorm/  $\rightarrow$  [gor\textipa{@}m] \\
         {\it gorm} `blue'
     \b. /banb\textipa{@}/ $\rightarrow$ [ban\textipa{@b@}] \\
         {\it Banba} `a name for Ireland'
     \b. /kon\textipa{f@}/ $\rightarrow$ [kon\textipa{@f@}] \\
         {\it confadh} `anger'
     \b. /an\textsuperscript{j}m\textsuperscript{j}/ $\rightarrow$ [an\textsuperscript{j}im\textsuperscript{j}] \\
         {\it ainm} `name'
     \z.
     (\citet{carnie.1994}:(37) and \citet{ni.chiosain.1999}:(2))

@@

\ex. \a. [\textipa{mna:}] or [\textipa{mra:}] (depending on dialect) \\
         {\it mn\'a} `women'
     \b. [\textipa{m\textsuperscript{j}r\textsuperscript{j}i:ra:n}] \\
         {\it mriath\'an} `sea-rods'
     \z.
     (\citet{swingle.1992}: 456)

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[table]{xcolor}    %for shading*
\usepackage{fullpage}
\usepackage{natbib}
\usepackage{mathpazo}
\usepackage{tipa}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{linguex}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows}
\usepackage{slashbox}
\usepackage{pifont}    %for pointing hand
\usepackage{arydshln}    %for dashed lines
\usepackage{rotating}    %for angled text

@@

\ex. \a. /alb\textipa{@}/ $\rightarrow$ [al\textipa{@b@}] \\
         {\it Alba} `Scotland'
     \b. /dorx\textipa{@}/ $\rightarrow$ [dor\textipa{@x@}] \\
         {\it dorcha} `dark'
     \b. /gorm/  $\rightarrow$  [gor\textipa{@}m] \\
         {\it gorm} `blue'
     \b. /banb\textipa{@}/ $\rightarrow$ [ban\textipa{@b@}] \\
         {\it Banba} `a name for Ireland'
     \b. /kon\textipa{f@}/ $\rightarrow$ [kon\textipa{@f@}] \\
         {\it confadh} `anger'
     \b. /an\textsuperscript{j}m\textsuperscript{j}/ $\rightarrow$ [an\textsuperscript{j}im\textsuperscript{j}] \\
         {\it ainm} `name'
     \z.
     \citep[(37)]{carnie.1994} and \citep[(2)]{ni.chiosain.1999}

@@

\ex. \a. [\textipa{mna:}] or [\textipa{mra:}] (depending on dialect) \\
         {\it mn\'a} `women'
     \b. [\textipa{m\textsuperscript{j}r\textsuperscript{j}i:ra:n}] \\
         {\it mriath\'an} `sea-rods'
     \z.
     \citep[456]{swingle.1992}

There are two classes of exceptions to this process.  First, homorganic clusters are not broken up by epenthesis.

\ex. \a. /farno\textipa{:g}/ $\rightarrow$ [farno\textipa{:g}], *[far\textipa{@}no\textipa{:g}] \\
         {\it fearnog} `alder'
     \b. /bord/ $\rightarrow$ [bord], *[bor\textipa{@}d] \\
         {\it bord} `table'
     \b. /kalr\textipa{@}/ $\rightarrow$ [kalr\textipa{@}], *[kal\textipa{@r@}] \\
         {\it calra} `calorie'  
     \z. 
     \citep[(52)]{carnie.1994}\footnote{\citet{ni.chiosain.1999} has a more complicated set of facts as to which homorganic clusters are broken up by epenthesis. As this is not the focus of this paper, I will assume the simpler dataset that Carnie presents.}

Second, epenthesis does not apply when the second consonant is a voiceless stop.

\ex. \a. /kork/ $\rightarrow$ [kork], *[kor\textipa{@}k] \\
         {\it Cork} `Cork (place name)' 
     \b. /kork\textipa{@}/ $\rightarrow$ [kork\textipa{@}], *[kor\textipa{@k@}] \\
         {\it corca} `people' 
     \b. /korp/  $\rightarrow$ [korp], *[kor\textipa{@p}] \\
         {\it corp} `body' %TODO : get some NT data in here!
     \z.
     \citep[(39a,b,c)]{carnie.1994}

\subsection{Analysis}

The fact that this process of epenthesis does not target obstruent-initial clusters is not problematic under the sonority contour faithfulness approach,
as the {\sc Ident(SonAngle)} constraints limit epenthesis, rather than trigger it.  By choosing a more limited markedness constraint such as {\sc *Son-C},
defined below, we can have epenthesis act only on sonorant-initial clusters.

\ex. {\sc *Son-C}: Assign a violation mark for every sonorant consonant preceding another consonant.

\ex. {\sc *Son-C} $\gg$ {\sc Dep}

\vspace{-2em}
\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{gorm}/} & {\sc *Son-C} & {\sc Dep} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{gor@m} & & \cellcolor{lightgray}$\ast$ \\
\hline b. & & \textipa{gorm} & $\ast$! & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

Epenthesis does not break up onset sonorant-initial clusters.  This is likely due to the fact that
with a small number of lexical exceptions, Irish words are stress-initial \citep[26]{o.siadhail.1989}.\footnote{With the exception of Munster Irish, where stress is attracted to syllables containing
a long vowel or diphthong (Green 1996), but where onset clusters still are not broken up by
epenthesis.  In this case, one would probably have to posit a positional faithfulness constraint
protecting the word-initial cluster.}  
Breaking up the onset cluster would result in an unstressable epenthetic vowel occupying the 
preferred location of stress in Irish.

I will assume initial stress is due to the following constraint, which is dominated only by 
lexicon-specific constraints.

\ex. {\sc Stress-Initial}: Assign a violation mark if the initial syllable of a word is not stressed.

\citet{alderete.2000} proposes a general constraint {\sc Head-Dependence} against making an epenthetic segment a prosodic head, which I adopt and state informally here:

\ex. {\sc Head-Dep}: Assign a violation mark for every stressed vowel not present in the input.

Together, these two constraints ranking above {\sc *Son-C} predict that onset epenthesis does not occur.

\ex. {\sc Stress-Initial}, {\sc Head-Dep} $\gg$ {\sc *Son-C}

\vspace{-2em}
\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c:c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{mna:}/} & {\sc Stress-Initial} & {\sc Head-Dep} & {\sc *Son-C} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{"mna:} & & & \cellcolor{lightgray}$\ast$ \\
\hline b. & & \textipa{"m@.na:} & & $\ast$! & \cellcolor{lightgray} \\
\hline c. & & \textipa{m@."na:} & $\ast$! & & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}
